---
title: "3D fluorescence microscopy – reproducibility analysis using mixed-effects
  models"
author: "Kanad Mandke, BABS STP"
date: "2026-01-06"
output:
  word_document: default
  html_document: default
  pdf_document: default
team: Hugh Sparks, Christopher Dunsby (Imperial)
---

The goal is to assess reproducibility of the experimental and analysis pipeline across four sites.

My assumptions are - 

- The underlying biological process is identical across sites.
- Each site ran one 24-well plate (6 conditions × 4 technical replicates).
- Reagents were independently purchased from the same supplier.
- No systematic batch effects are assumed a priori.
- 10 spheroids randomly selected per well, image them in 3D, segment cells, and quantify fluorescence per cell.
- Cells within spheroids are subject to segmentation noise. Note that cells within spheroids are more similar to each other than between spheroids. 
- Therefore, inference is performed at the spheroid level, not at the cell level.

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries
```{r}

library(tidyverse)
library(lme4)
library(lmerTest)
library(dplyr)
library(readr)
library(writexl)
library(ggeffects)
library(ggplot2)
library(ggrain)
library(emmeans)
library(broom.mixed)
library(tibble)
library(performance)

```

# Load data
```{r}
# Import data and perform some data checks
df <- read_csv("/nemo/stp/babs/working/mandkek/projects/dunsbycw/sparks.hugh/data/master_cross_partner.csv")
glimpse(df)

# data checks
df %>% count(Site)
df %>% count(Site, Condition)
df %>% count(Site, well)

```

# Aggregate at spheroid level - DV: cyt_nuc_ratio_sensor
```{r} 

df_spheroid <- df %>% group_by(Site, Condition, Spheroid, well) %>%
  summarise(mean_cnr = mean(cyt_nuc_ratio_sensor, na.rm = TRUE), 
            n_cells = n(),
            .groups = "drop")

glimpse(df_spheroid)


# Explicitly order Condition factor, intercept/reference = DMSO
df_spheroid <- df_spheroid %>%
  mutate(Condition = factor(Condition,
        levels = c(
        "DMSO",
        "20 nM Binimetinib",
        "60 nM Binimetinib",
        "200 nM Binimetinib",
        "100 nM TPA +ve control",
        "3.5 µM Binimetinib -ve control")))

# Sanity check
levels(df_spheroid$Condition)


```

# Data vis - ratio (raincloud plots)
```{r}

ggplot(df_spheroid, aes(x = Condition, y = mean_cnr, fill = Condition)) +
  geom_rain(alpha = 0.6) + facet_wrap(~ Site) + theme_classic() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1, vjust = 1), # change angle values for readability 
        legend.position = "none") +
  labs(x = "Condition",y = "Mean cyt:nuc ratio")


```

# Hierarchical mixed-effects model 
DV: cytoplasm to nucleus ratio
Fixed effects: Condition
Random effects: Site
              : Well nested within site (technical replicates)
```{r}

m_mixed <- lmer(mean_cnr ~ Condition + (1 | Site) + (1 | Site:well),
                data = df_spheroid,REML = FALSE) # REML = FALSE forces maximum likelihood

summary(m_mixed)
icc(m_mixed) # what proportion of variance is due to random groupings (of site)

# QQ plot - diagnostic 
qqnorm(resid(m_mixed)) 
qqline(resid(m_mixed), col = "red")

```

# Data vis - mixed effects model
```{r}

# Create a simple plot of marginal means from mixed model
emm_mixed <- ggpredict(m_mixed, terms = "Condition")

ggplot() + geom_jitter(data = df_spheroid,
        aes(x = Condition, y = mean_cnr), width = 0.15, alpha = 0.5, colour = "grey40") +
  geom_point(data = emm_mixed,
        aes(x = x, y = predicted), size = 3,colour = "black") +
  geom_errorbar(data = emm_mixed, aes(x = x, ymin = conf.low, ymax = conf.high), width = 0.2) +
  theme_classic() +  theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
  labs(x = "Condition", y = "Mean cyt:nuc ratio",
    title = "Mixed-effects model: marginal means (±95% CI)")

```


# Site as fixed effects - this is for contrasts
```{r}

# Reference site is Crick - alphabetical 
m_fixed <- lmer(mean_cnr ~ Condition + Site + (1 | Site:well),
                data = df_spheroid,REML = FALSE)

summary(m_fixed)
icc(m_fixed)

# Extract site contrasts
tidy(m_fixed, effects = "fixed") %>% filter(str_detect(term, "Site"))

qqnorm(resid(m_fixed))
qqline(resid(m_fixed), col = "red")


```


# Data vis - site as fixed effects
```{r}

emm_fixed <- ggpredict(m_fixed, terms = c("Condition", "Site"))

# Reference site is Crick - alphabetical 
ggplot(emm_fixed, aes(x = x, y = predicted, colour = group)) +
  geom_point(position = position_dodge(width = 0.3), size = 2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2,position = position_dodge(width = 0.3)) +
  theme_classic() + theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
  labs(x = "Condition", y = "Mean cyt:nuc ratio", colour = "Site",
    title = "Fixed-effects model: condition effects by site")



```

# Run a full model
```{r}

# Now that we have established that there are site specific differences we run a full model
m_full <- lmer(mean_cnr ~ Condition * Site + (1 | Site:well),
     data = df_spheroid, REML = TRUE)

summary(m_full)
print(m_full, correlation=TRUE) # print correlations

qqnorm(resid(m_full))
qqline(resid(m_full), col = "red")

```

# Data vis - full model
```{r}

emm <- emmeans(m_full, ~ Condition | Site)

emm_df <- as.data.frame(emm)

ggplot(emm_df, aes(x = Condition, y = emmean, color = Site, group = Site)) +
  geom_point(position = position_dodge(width = 0.3), size = 3) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE),width = 0.15,position = position_dodge(width = 0.3)) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1)) +
  labs(
    y = "Model-estimated mean cyt:nuc ratio",
    x = "Condition",
    title = "Condition × Site effects from mixed-effects model",
    subtitle = "Points and intervals represent model-adjusted estimates")

# Variability is quite small (this is good!)
```

# Save model outputs
```{r}

# All fixed effects
fixed_effects <- function(model, model_name) 
  {tidy(model, effects = "fixed") %>%
    mutate(model = model_name)}

fixed_effects_tbl <- bind_rows(fixed_effects(m_mixed, "m_mixed"),
fixed_effects(m_fixed, "m_fixed"), fixed_effects(m_full,  "m_full"))

# --------------------------------------------- #
# All random effects
random_effects <- function(model, model_name) 
  {tidy(model, effects = "ran_pars") %>%
    mutate(model = model_name)}

random_effects_tbl <- bind_rows(random_effects(m_mixed, "m_mixed"),
random_effects(m_fixed, "m_fixed"),random_effects(m_full,  "m_full"))

# --------------------------------------------- #
# Model fit stats
model_stats <- function(model, model_name)
  {glance(model) %>%
  mutate(model = model_name)}

fit_stats_tbl <- bind_rows(model_stats(m_mixed, "m_mixed"),
model_stats(m_fixed, "m_fixed"),model_stats(m_full,  "m_full"))

# ---------------------------------------------- #
# Fixed effects correlations
fixed_effect_correlations <- function(model, model_name) 
{cor_mat <- cov2cor(as.matrix(vcov(model)))
as.data.frame(cor_mat) %>%
rownames_to_column("term_1") %>%
pivot_longer(-term_1,names_to = "term_2",values_to = "correlation") %>%
mutate(model = model_name)}

correlations_tbl <- bind_rows(fixed_effect_correlations(m_mixed, "m_mixed"),
fixed_effect_correlations(m_fixed, "m_fixed"),fixed_effect_correlations(m_full,  "m_full"))

# ----------------------------------------- #
# Write excel
write_xlsx(
  list(
    "Fixed effects"        = fixed_effects_tbl,
    "Random effects"       = random_effects_tbl,
    "Model fit statistics" = fit_stats_tbl,
    "Fixed-effect correlations" = correlations_tbl),
  path = "results/lmer_model_outputs.xlsx") # Set path

```
